Simulation Learning in Computer Hex

### 2011-08-23 thesis/thesis.tex
A bit more in every section plus more references.

\chapter{Introduction}
Abstract strategy games provide an easy testbed for artificial intelligence and machine learning applications. The game framework provides immediate feedback on algorithm performance. The choice of game and rules can also be experimentally chosen for a variety of difficulty and computational complexity levels. Starting with mathematical game theory research, the most common technique for game AI has been brute force search of the tree of legal moves. This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World-Champion Gary Kasparov. These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for hundereds of moves. Kasparov himself suggested a version of Chess that allowed the player to consult a computer for move advice. Chinook went so far as to solve checkers, proving that perfect play will end in a draw. For at least Checkers, Chess, and similar games (Othello, Connect Four), brute force tree search has provided a silver bullet.

@@

@@

\chapter{State of the Art in Monte Carlo Tree Search}\label{mcts}
As stated, the current top computer Hex player, MoHex, uses the open-source Fuego implementation of Monte-Carlo Tree Search (MCTS). The basic Monte-Carlo technique for Computer Go was proposed by Bru\"{u}gmann\cite{brugmann1993monte}. His program \emph{Mango} effectively used a ``flat'' search tree (1-ply) and simply ran simulations from each candidate move to estimate its value. To contrast, MCTS uses Monte-Carlo simulations to grow a minimax search tree. The UCT algorithm was developed alongside MCTS to direct the growth of the search tree\cite{gelly2006exploration}. At any point, the search can be stopped and the current state of the tree used to select a move to play. Performance of the search is primarily based on three factors: the \emph{selection policy}, the \emph{simulation policy}, and the \emph{update policy}.

@@

\section{Biasing the Simulation}\label{bias}
Gelly and others have shown that using heuristics in both the expansion and playout policy can prove extremely beneficial\cite{gelly2006modification}\cite{gelly2008achieving}. However, this frequently requires expert knowledge (heuristics) that can be hard to obtain or not available for some games (particularly new games). It was also found that improving the performance of playout policy did not guarantee an improvement in the tree search. For example, the playout policy can implicitly prune good moves, never allowing them to be played. In Go, the policy must correctly handle complicated situations like Nakade. If it fails to correctly play out a complex position, the tree might fail to see an opponent response. Much time and energy has been spent on hand-tuning the policies of the top engines to correctly handle complicated tactical situation. If there is some small probability of playing a purely random move, the convergence properties of the UCT algorithm will still hold, but can require a prohibitive amount of playouts to ``find'' the move sequence. In fact, one of the top Go engines, MoGo, uses a near-zero exploration coefficient in the selection policy\cite{gelly2007combining}. This puts most of the control of the search in the hands of the simulation policy.

\chapter{Problem Statement}\label{problem}

\chapter{Solution}\label{solution}
After each move, a random move is selected in the 6-connected local area. This move is used as the next move in the playout. All subsequent moves will also have their local area examined for a random response. If all 6 connected vertices are non-empty, a move is played using the uniform random policy (the \emph{default} policy). In the uniform local variation, a response is chosen uniformly from any empty moves in the local area. The uniform local variation with tenuki is similar, but there is a probability $p=\frac{1}{6}$ of using the default policy. The learned local variation assigns a weight to each empty local vertex based on the local area. These weights are used to stochastically bias the local response. The weights are trained over time through self-play using Evolution Strategies.

\section{Evolution Strategies}
Evolution strategies (ES) is a collection of algorithms that operate on real-valued vectors. A general form of ES is given here, with some small modifications required for it to represent a biased simulation policy discussed in Section \ref{esmods}. More information on evolution strategies and evolutionary learning in general can be found in Eberhard and Shi's excellent textbook \emph{Computational Intelligence: Concepts to Implementations}\cite{eberhart2007computational}. The Evolution Strategies article on Scholarpedia.org also provides a good introduction to the topic\cite{Beyer:2007}.

@@

\section{Using ES to Bias the Simulation}\label{esmods}

\section{Fitness Evaluation}\label{fitness}

\chapter{Results and Conclusions}\label{results}
To test the effects of learning policy weights, the evolution process was ran for 41 generations. This took approximately 2 days using a single core of an Intel Xeon CPU running at 2.80GHz. Although evolution strategies is easy to parallelize by running multiple evaluations simultaneously, the current implementation is single-threaded. After training was completed, the weights of top 4 individuals of the 41st generation were blended in proportion to their fitness to form a single individual. This final individual was used to bias the simulation policy and compared to the uniform local bias, uniform local bias with a $\frac{5}{6}$th probability of playing locally. The control consisted of uniform random playouts. All players used a slightly modified UCT selection policy that included the variance term suggested by Gelly 2006\cite{gelly2006exploration}. A pure update policy was used, only the nodes directly in the path traversed by UCT were updated with the result of a simulation.

>>>>

\renewcommand*\thesection{\arabic{section}}

\section{Introduction}
Abstract strategy games provide an easy testbed for artificial intelligence and machine learning applications. The game framework provides immediate feedback on algorithm performance. The choice of game and rules can also be experimentally chosen for a variety of difficulty and computational complexity levels. Starting with mathematical game theory research, the most common technique for game AI has been brute force search of the tree of legal moves. This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World-Champion Gary Kasparov. These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for hundereds of moves. Kasparov himself suggested a version of Chess that allowed the player to consult a computer for move advice. Chinook went so far as to solve checkers, proving that perfect play will end in a draw. For at least Checkers, Chess, and similar games (Othello, Connect Four), brute force tree search has provided a silver bullet.

@@

Hex is very similar to the game of Havannah, another ``connection'' game. It also has similarities to the board game Go (also known as Igo, Weiqi, and Baduk). Go is a much older game than Hex or Havannah, and much research has been done on Go playing algorithms. Hex has simpler rules than Go, but has a similar branching factor and complexity. Much of this thesis will use Hex as a simpler analog for Go, with the techniques discussed being portable for use in any large branching game without easily crafted heuristics.

@@

\section{State of the Art in Monte Carlo Tree Search}\label{mcts}
As stated, the current top computer Hex player, MoHex, uses the open-source Fuego implementation of Monte-Carlo Tree Search (MCTS). The basic Monte-Carlo technique for Computer Go was proposed by Bru\"{u}gmann\cite{brugmann1993monte}. His program \emph{Mango} effectively used a ``flat'' search tree (1-ply) and simply ran simulations from each candidate move to estimate its value. To contrast, MCTS uses Monte-Carlo simulations to grow a minimax search tree. The UCT algorithm was developed alongside MCTS to direct the growth of the search tree\cite{gelly2006exploration}. At any point, the search can be stopped and the current state of the tree used to select a move to play. Performance of the search is primarily based on three factors: the \emph{selection policy}, the \emph{simulation policy}, and the \emph{update policy}.

@@

\subsection{Biasing the Simulation}\label{bias}
Gelly and others have shown that using heuristics in both the expansion and playout policy can prove extremely beneficial\cite{gelly2006modification}\cite{gelly2008achieving}. However, this frequently requires expert knowledge (heuristics) that can be hard to obtain or not available for some games (particularly new games). It was also found that improving the performance of playout policy did not guarantee an improvement in the tree search. For example, the playout policy can implicitly prune good moves, never allowing them to be played. In Go, the policy must correctly handle complicated situations like Nakade. If it fails to correctly play out a complex position, the tree might fail to see an opponent response. Much time and energy has been spent on hand-tuning the policies of the top engines to correctly handle complicated tactical situation. If there is some small probability of playing a purely random move, the convergence properties of the UCT algorithm will still hold, but can require a prohibitive amount of playouts to ``find'' the move sequence. In fact, one of the top Go engines, MoGo, uses a near-zero exploration coefficient in the selection policy\cite{gelly2007combining}. This puts most of the control of the search in the hands of the simulation policy.

\section{Problem Statement}\label{problem}
MoGo, and Go programs in general, have the benefit of a rich history of play. Thousands of expert games are available for analysis. Some approaches to improving the simulation policy have focused on using expert games in training weights to direct the playout\cite{chaslot2010adding}. More generally, common patterns are used in a hand-crafted policy to select ``urgent'' moves to play in a local area around the last played move. Hex does have one simple pattern in the form of a bridge\cite{anshelevich2002hierarchical}, but few easy paths to crafting a better simulation policy.

It would be ideal to automatically learn a better playout policy from self-play. The usual process of improvement requires hand-tuned weights that modify the playout policy. Any change must be tested to detect the change in strength it gives. Problematically, small changes in overall algorithm strength require thousands of games of testing to detect. Many times the improvement made by a small change in weight can be so small is will go unnoticed. Automatic weight tuning can improve this process, but still requires thousands of games to test. Recent work in this area has focused on learning weights for a small number of patterns using expert knowledge and reinforcement learning\cite{silver2009monte}. This thesis presents a slightly different approach using evolutionary learning. It solves the problem of requiring a strong outside player or expert knowledge by purely using games of self-play.

\section{Solution}\label{solution}
After each move, a random move is selected in the 6-connected local area. This move is used as the next move in the playout. All subsequent moves will also have their local area examined for a random response. If all 6 connected vertices are non-empty, a move is played using the uniform random policy (the \emph{default} policy). In the uniform local variation, a response is chosen uniformly from any empty moves in the local area. The uniform local variation with tenuki is similar, but there is a probability $p=\frac{1}{6}$ of using the default policy. The learned local variation assigns a weight to each empty local vertex based on the local area. These weights are used to stochastically bias the local response. The weights are trained over time through self-play using Evolution Strategies.

\subsection{Evolution Strategies}
Evolution strategies (ES) is a collection of algorithms that operate on real-valued vectors. A general form of ES is given here, with some small modifications required for it to represent a biased simulation policy discussed in Section \ref{esmods}. More information on evolution strategies and evolutionary learning in general can be found in Eberhard and Shi's excellent textbook \emph{Computational Intelligence: Concepts to Implementations}\cite{eberhart2007computational}. The Evolution Strategies article on Scholarpedia.org also provides a good introduction to the topic\cite{Beyer:2007}.

@@

\subsection{Using ES to Bias the Simulation}\label{esmods}

\subsection{Fitness Evaluation}\label{fitness}
Every generation, each individual plays $n$ games against another random individual. Roughly half the games are played as black, half as white. The swap-safe rule was used to play the first move to further equalize the first-mover advantage. The winner of a game receives 1 point added to their fitness, the loser has 1 point deducted from their fitness.

\section{Results and Conclusions}\label{results}
To test the effects of learning policy weights, the evolution process was ran for 41 generations. This took approximately 2 days using a single core of an Intel Xeon CPU running at 2.80GHz. Although evolution strategies is easy to parallelize by running multiple evaluations simultaneously, the current implementation is single-threaded. After training was completed, the weights of top 4 individuals of the 41st generation were blended in proportion to their fitness to form a single individual. This final individual was used to bias the simulation policy and compared to the uniform local bias, uniform local bias with a $\frac{5}{6}$th probability of playing locally. The control consisted of uniform random playouts. All players used a slightly modified UCT selection policy that included the variance term suggested by Gelly 2006\cite{gelly2006exploration}. A pure update policy was used, only the nodes directly in the path traversed by UCT were updated with the result of a simulation.
